Solar wind predictions for the Parker Solar Probe orbit

### 2017-10-22 paperMVVB/sections/content.tex
#2 integrated changes by MVVB

%\subsection{SSN correlation}
For the correlations we fit lognormal functions to the frequency distributions as in Sect.~\ref{sec:frequency_distribution}, but implement linear relations to the yearly SSN, allowing shifting of the distribution functions with SSN. For the velocity the approach is different insofar as its two components are kept fixed and instead their balance is modified with the changing SSN. \textbf{Thus we obtain on solar activity depending models for the frequency distributions of all four solar wind parameters.}

@@

%\subsection{SSN fitting}
\textbf{Next we create solar activity dependent analytical representations of the solar wind frequency distributions. This is achieved by shifting the median positions of the lognormal distributions as a linear function of SSN. To enable these shifts }, we add a linear SSN dependency to the median
\begin{align}
	x_\text{med}(ssn) &= a_\text{med} \cdot ssn + b_\text{med}\,,	\label{eq:median_with_ssn}
\end{align}
using a factor to the SSN $a_\text{med}$ with a baseline $b_\text{med}$. We relate the mean with a scaling factor to the median to transfer its SSN dependency:
\begin{align}
	x_\text{avg}(ssn) &= (1 + a_\text{avg}) \cdot x_\text{med}(ssn)\,.	\label{eq:mean_with_ssn}
\end{align}
\textbf{These relations, substituted into the lognormal function (\ref{eq:single_lognormal_fit_function}), lead to a new SSN-dependent function $W'(x,ssn)$. This function} is then fitted to the yearly data\textbf{, using the yearly SSN as input parameter.} \textbf{The SSN is offset with the individual time lags for each parameter determined before, to profit from the higher correlation.} The three resulting fit coefficients ($a_\text{med}, b_\text{med}$ and $a_\text{avg}$) are presented in Table~\ref{tab:ssn_fit_parameters}.
\begin{table*}
	\caption{Resulting fit coefficients from the OMNI data fitting with \textbf{the linear SSN dependencies (\ref{eq:median_with_ssn}) and (\ref{eq:mean_with_ssn})}. For the velocity the fit parameters from the double lognormal fit \textbf{(\ref{eq:double_lognormal_fit_function})} and their balancing function \textbf{(\ref{eq:balance_with_ssn})} are given. \textbf{The numbers in parentheses are the estimated standard deviations of the fit parameters referred to the corresponding last digits of the quoted value. The listed SSN time lags are used for the fits.}}
	\label{tab:ssn_fit_parameters}
	\centering
	\begin{tabular}{l@{} c@{}
		S[table-format = 1.3(2)e+1]
		S[table-format = 1.4(2)]
		S[table-format = 1.3(2)e+1]
		S[table-format = +1.3(2)e+1]
		c c
		}
		\hline\hline
		\multicolumn{2}{l}{\multirow{2}{*}{Parameter}}	&\multicolumn{2}{c}{Median}	&\multicolumn{1}{c}{Mean}	&\multicolumn{2}{c}{Balance}	&\multicolumn{1}{c}{SSN lag}\\
		\cline{3-4}\cline{6-7}
		\multicolumn{2}{l}{}	&\multicolumn{1}{c}{SSN factor $a_\text{med}$}	&\multicolumn{1}{c}{Baseline $b_\text{med}$}	&\multicolumn{1}{c}{Scaling factor $a_\text{avg}$}	&\multicolumn{1}{c}{SSN factor $c_a$}	&\multicolumn{1}{c}{Baseline $c_b$}	&\multicolumn{1}{c}{[years]}\\
		\hline
		\multicolumn{2}{l}{\textbf{Magnetic field [\si{nT}]}}	&1.309(19)e-2	&4.285(17)	&8.786(78)e-2	&\multicolumn{1}{c}{--}	&--	&0\\
%		\multicolumn{2}{l}{Velocity}	&2.514(93)e-3	&3.8521(90)	&2.392(26)e-2	&\multicolumn{1}{c}{--}	&--\\
		\multicolumn{2}{l}{\textbf{Density [\si{\per\cm\cubed}]}}	&3.81(25)e-3	&4.495(26)	&3.050(27)e-1	&\multicolumn{1}{c}{--}	&--	&6\\
		\multicolumn{2}{l}{\textbf{Temperature [\SI{e4}{\K}]}}	&1.974(26)e-2	&5.729(19)	&6.541(28)e-1	&\multicolumn{1}{c}{--}	&--	&3\\
		\hline
		\multicolumn{1}{l}{\textbf{Velocity}}	&\multicolumn{1}{c}{$W'_1$}	&\multicolumn{1}{c}{--}	&3.633(12)	&1.008(37)e-2	&\multirow{2}{*}{\tablenum{-1.799(95)e-3}}	&\multirow{2}{*}{0.638(32)}	&\multirow{2}{*}{3}\\
		\multicolumn{1}{l}{\textbf{[\SI{e2}{\km\per\s}]}}	&\multicolumn{1}{c}{$W'_2$}	&\multicolumn{1}{c}{--}	&4.831(81)	&2.31(20)e-2	&	&	&\\
		\hline
	\end{tabular}
\end{table*}

@@

\textbf{The radial dependencies of the key solar wind parameters over the distance range \SIrange{0.29}{0.98}{\au} measured by both Helios probes are plotted in Fig.~\ref{fig:radial_fit_4_thesis_light_b_skip}, together with their median and mean values for different solar distances, calculated for the minimal distance resolution \SI{0.01}{\au} of the data set.}
\begin{figure*}
	\includegraphics[width=18cm]{figures/radial_fit_4_thesis_light_b_skip.pdf}
	\caption{Helios hourly data plots of the four solar wind parameters over solar distance. The mean and median per \SI{0.01}{\au} data bin and their fit curves are plotted as well. The Helios data has a native distance resolution of \SI{0.01}{\au}, thus, to make the abundance visible in these plots, we added a random distance value of up to \SI{+-0.005}{\au}. The maximal velocities of CME events larger than \SI{800}{\km\per\s} are indicated with circles.}
	\label{fig:radial_fit_4_thesis_light_b_skip}
\end{figure*}
Assuming a radial solar wind outflow, it is expected that the distance dependence of the solar wind parameters over the Helios data range \SIrange{0.29}{0.98}{\au} can be described through power law scaling. Therefore we use the power law function
\begin{align}
	x(r) = d\cdot r^e	\label{eq:power_function}
\end{align}
for the regression fit of the median and mean, with $r$ being the solar distance in astronomical units, $d$ the magnitude at \SI{1}{\au} and $e$ the exponent. The fits are weighted through the different data counts per bin.
%fit result table and figure
\textbf{The obtained coefficients for the median and mean power law fits ($d_\text{med}$, $e_\text{med}$, $d_\text{avg}$ and $e_\text{avg}$) are listed in Table~\ref{tab:mean_median_fit_parameter} and their corresponding curves are shown in Fig.~\ref{fig:radial_fit_4_thesis_light_b_skip}.}
\begin{table*}
	\caption{Fit coefficients for the median and mean solar distance dependencies \textbf{(\ref{eq:power_function})} of the four solar wind parameters derived from the combined Helios~1 and 2 data. \textbf{The numbers in parentheses are the estimated standard deviations of the fit parameters referred to the corresponding last digits of the quoted value.} The crossing distances indicate where the median and mean fits intersect each other. The yearly variation is the weighted standard deviation derived from the yearly fit exponents seen in Fig.~\ref{fig:yearly_gradients_c}.}
	\label{tab:mean_median_fit_parameter}
	\centering
	\begin{tabular}{l
	S[table-format = 1.3(2)]
	S[table-format = +1.3(2)]
	@{}c
	S[table-format = 1.3(2)]
	S[table-format = +1.3(2)]
	S[table-format = 1.1(2)e1]
	S[table-format = 1.3]}
		\hline\hline
		\multirow{2}{*}{Parameter}	&\multicolumn{2}{c}{Median}	&	&\multicolumn{2}{c}{Mean}	&\multicolumn{1}{c}{Crossing distance}	&\multicolumn{1}{c}{Yearly variation}\\
		\cline{2-3}	\cline{5-6}
			&\multicolumn{1}{c}{$d_\text{med}$}	&\multicolumn{1}{c}{$e_\text{med}$}	&	&\multicolumn{1}{c}{$d_\text{avg}$}	&\multicolumn{1}{c}{$e_\text{avg}$}	&\multicolumn{1}{c}{[\si{\au}]}	&\multicolumn{1}{c}{$\Delta e$}\\
		\hline
		\textbf{Magnetic field [\si{nT}]}	&5.377(92)	&-1.655(17)	&	&6.05(10)	&-1.546(18)	&0.339(11)	&0.11\\
		\textbf{Velocity [\SI{e2}{\km\per\s}]}	&4.107(28)	&0.058(13)	&	&4.356(24)	&0.049(10)	&0.7(83)e3	&0.012\\
		\textbf{Density [\si{\per\cm\cubed}]}	&5.61(27)	&-2.093(46)	&	&7.57(30)	&-2.010(38)	&0.027(73)	&0.072\\
		\textbf{Temperature [\SI{e4}{\K}]}	&7.14(23)	&-0.913(39)	&	&9.67(21)	&-0.792(28)	&0.082(85)	&0.050\\
		\hline
	\end{tabular}
\end{table*}
% source year variation: fit_out_yearly_gradients_table_b_stats.txt

@@

%crossing distances
\textbf{Our from the Helios data fit acquired} mean and median velocity fit exponents are very similar, which indicates that they just as well can be kept identical so that the basic shape of the frequency distribution does not change with distance. Contrary, the mean and median fits for the magnetic field strength cross each other at \SI{0.339}{\au} \textbf{(see Table~\ref{tab:mean_median_fit_parameter})} and the mean \hl{is \textbf{getting slightly} lower than the median at smaller distances }. Thus, below that distance the \textbf{frequency distribution} cannot well be described anymore by a lognormal function\textbf{, because the mean of a lognormal function has to be larger than its median (as mentioned in Sect.~\ref{sec:frequency_distribution}), that is, the location of the crossing indicates that the parameter's distribution is not anymore of a lognormal shape thereafter.} The fits for the proton temperature show a similar behavior, having an \textbf{extrapolated} intersection at \SI{0.082}{\au}. Therefore the extrapolation of the magnetic field and temperature distribution frequencies to the PSP orbit by applying lognormal functions is limited. \textbf{The crossing points limit the regions where the distribution's shapes can still be considered lognormal.}

\textbf{For being able to still fit and extrapolate lognormal functions with the data, we assume that the shapes keep being lognormal at all distances. For the coming frequency distribution fit function we reduce the fit exponents $e_\text{med}$ and $e_\text{avg}$ to only one.} It should be noted that this simplification leads to slightly larger modeling errors, especially in case of the magnetic field strength.

>>>>

%\subsection{SSN correlation}
For the correlations we fit lognormal functions to the frequency distributions as in Sect.~\ref{sec:frequency_distribution}, but implement linear relations to the yearly SSN, allowing shifting of the distribution functions with SSN. For the velocity the approach is different insofar as its two components are kept fixed and instead their balance is modified with the changing SSN. \textbf{Thus we obtain solar activity depending models for the frequency distributions of all four solar wind parameters.}

@@

%\subsection{SSN fitting}
\textbf{Next we create solar activity dependent analytical representations of the solar wind frequency distributions. This is achieved by shifting the median positions of the lognormal distributions as a linear function of SSN. To enable these shifts }, we add a linear SSN dependency to the median
\begin{align}
	x_\text{med}(ssn) &= a_\text{med} \cdot ssn + b_\text{med}\,,	\label{eq:median_with_ssn}
\end{align}
using a factor to the SSN $a_\text{med}$ with a baseline $b_\text{med}$. We relate the mean with a scaling factor to the median to transfer its SSN dependency:
\begin{align}
	x_\text{avg}(ssn) &= (1 + a_\text{avg}) \cdot x_\text{med}(ssn)\,.	\label{eq:mean_with_ssn}
\end{align}
\textbf{These relations, substituted into the lognormal function (\ref{eq:single_lognormal_fit_function}), lead to a new SSN-dependent function $W'(x,ssn)$. This function} is then fitted to the yearly data\textbf{, using the yearly SSN as input parameter.} \textbf{The SSN is offset with the individual time lags for each parameter determined before, to benefit from the higher correlation.} The three resulting fit coefficients ($a_\text{med}, b_\text{med}$ and $a_\text{avg}$) are presented in Table~\ref{tab:ssn_fit_parameters}.
\begin{table*}
	\caption{Resulting fit coefficients from the OMNI data \textbf{based on the linear SSN dependencies (\ref{eq:median_with_ssn}) and (\ref{eq:mean_with_ssn})}. For the velocity the fit parameters from the double lognormal fit \textbf{(\ref{eq:double_lognormal_fit_function})} and their balancing function \textbf{(\ref{eq:balance_with_ssn})} are given. \textbf{The numbers in parentheses are the estimated standard deviations of the fit parameters referred to the corresponding last digits of the quoted value. The listed SSN time lags are used for the fits.}}
	\label{tab:ssn_fit_parameters}
	\centering
	\begin{tabular}{l@{} c@{}
		S[table-format = 1.3(2)e+1]
		S[table-format = 1.4(2)]
		S[table-format = 1.3(2)e+1]
		S[table-format = +1.3(2)e+1]
		c c
		}
		\hline\hline
		\multicolumn{2}{l}{\multirow{2}{*}{Parameter}}	&\multicolumn{2}{c}{Median}	&\multicolumn{1}{c}{Mean}	&\multicolumn{2}{c}{Balance}	&\multicolumn{1}{c}{SSN lag}\\
		\cline{3-4}\cline{6-7}
		\multicolumn{2}{l}{}	&\multicolumn{1}{c}{SSN factor $a_\text{med}$}	&\multicolumn{1}{c}{Baseline $b_\text{med}$}	&\multicolumn{1}{c}{Scaling factor $a_\text{avg}$}	&\multicolumn{1}{c}{SSN factor $c_a$}	&\multicolumn{1}{c}{Baseline $c_b$}	&\multicolumn{1}{c}{[years]}\\
		\hline
		\multicolumn{2}{l}{\textbf{Magnetic field [\si{nT}]}}	&1.309(19)e-2	&4.285(17)	&8.786(78)e-2	&\multicolumn{1}{c}{--}	&--	&0\\
%		\multicolumn{2}{l}{Velocity}	&2.514(93)e-3	&3.8521(90)	&2.392(26)e-2	&\multicolumn{1}{c}{--}	&--\\
		\multicolumn{2}{l}{\textbf{Density [\si{\per\cm\cubed}]}}	&3.81(25)e-3	&4.495(26)	&3.050(27)e-1	&\multicolumn{1}{c}{--}	&--	&6\\
		\multicolumn{2}{l}{\textbf{Temperature [\SI{e4}{\K}]}}	&1.974(26)e-2	&5.729(19)	&6.541(28)e-1	&\multicolumn{1}{c}{--}	&--	&3\\
		\hline
		\multicolumn{1}{l}{\textbf{Velocity}}	&\multicolumn{1}{c}{$W'_1$}	&\multicolumn{1}{c}{--}	&3.633(12)	&1.008(37)e-2	&\multirow{2}{*}{\tablenum{-1.799(95)e-3}}	&\multirow{2}{*}{0.638(32)}	&\multirow{2}{*}{3}\\
		\multicolumn{1}{l}{\textbf{[\SI{e2}{\km\per\s}]}}	&\multicolumn{1}{c}{$W'_2$}	&\multicolumn{1}{c}{--}	&4.831(81)	&2.31(20)e-2	&	&	&\\
		\hline
	\end{tabular}
\end{table*}

@@

\textbf{The radial dependencies of the key solar wind parameters over the distance range \SIrange{0.29}{0.98}{\au} measured by both Helios probes are plotted in Fig.~\ref{fig:radial_fit_4_thesis_light_b_skip}, together with their median and mean values for different solar distances, calculated for the minimal distance resolution \SI{0.01}{\au} of the data set.}
\begin{figure*}
	\includegraphics[width=18cm]{figures/radial_fit_4_thesis_light_b_skip.pdf}
	\caption{Helios hourly data plots of the four solar wind parameters over solar distance. The mean and median per \SI{0.01}{\au} data bin and their fit curves are plotted as well. The Helios data has a native distance resolution of \SI{0.01}{\au}, thus, to make the abundance visible in these plots, we added a random distance value of up to \SI{+-0.005}{\au}. The higher velocity data points above \SI{800}{\km\per\s} \hl{identified as CME events by BS1998, Sheeley1981 are circled red.}}
	\label{fig:radial_fit_4_thesis_light_b_skip}
\end{figure*}
Assuming a radial solar wind outflow, it is expected that the distance dependence of the solar wind parameters over the Helios data range \SIrange{0.29}{0.98}{\au} can be described through power law scaling. Therefore we use the power law function
\begin{align}
	x(r) = d\cdot r^e	\label{eq:power_function}
\end{align}
for the regression fit of the median and mean, with $r$ being the solar distance in astronomical units, $d$ the magnitude at \SI{1}{\au} and $e$ the exponent. The fits are weighted through the different data counts per bin.
%fit result table and figure
\textbf{The obtained coefficients for the median and mean power law fits ($d_\text{med}$, $e_\text{med}$, $d_\text{avg}$ and $e_\text{avg}$) are listed in Table~\ref{tab:mean_median_fit_parameter} and their corresponding curves are shown in Fig.~\ref{fig:radial_fit_4_thesis_light_b_skip}.}
\begin{table*}
	\caption{Fit coefficients for the median and mean solar distance dependencies \textbf{(\ref{eq:power_function})} of the four solar wind parameters derived from the combined Helios~1 and 2 data. \textbf{The numbers in parentheses are the estimated standard deviations of the fit parameters referred to the corresponding last digits of the quoted value.} The crossing distances indicate where the median and mean fits intersect each other. The yearly variation is the weighted standard deviation derived from the yearly fit exponents seen in Fig.~\ref{fig:yearly_gradients_c}.}
	\label{tab:mean_median_fit_parameter}
	\centering
	\begin{tabular}{l
	S[table-format = 1.3(2)]
	S[table-format = +1.3(2)]
	@{}c
	S[table-format = 1.3(2)]
	S[table-format = +1.3(2)]
	S[table-format = 1.1(2)e1]
	S[table-format = 1.3]}
		\hline\hline
		\multirow{2}{*}{Parameter}	&\multicolumn{2}{c}{Median}	&	&\multicolumn{2}{c}{Mean}	&\multicolumn{1}{c}{Crossing distance}	&\multicolumn{1}{c}{Yearly variation}\\
		\cline{2-3}	\cline{5-6}
			&\multicolumn{1}{c}{$d_\text{med}$}	&\multicolumn{1}{c}{$e_\text{med}$}	&	&\multicolumn{1}{c}{$d_\text{avg}$}	&\multicolumn{1}{c}{$e_\text{avg}$}	&\multicolumn{1}{c}{[\si{\au}]}	&\multicolumn{1}{c}{$\Delta e$}\\
		\hline
		\textbf{Magnetic field [\si{nT}]}	&5.377(92)	&-1.655(17)	&	&6.05(10)	&-1.546(18)	&0.339(11)	&0.11\\
		\textbf{Velocity [\SI{e2}{\km\per\s}]}	&4.107(28)	&0.058(13)	&	&4.356(24)	&0.049(10)	&0.7(83)e3	&0.012\\
		\textbf{Density [\si{\per\cm\cubed}]}	&5.61(27)	&-2.093(46)	&	&7.57(30)	&-2.010(38)	&0.027(73)	&0.072\\
		\textbf{Temperature [\SI{e4}{\K}]}	&7.14(23)	&-0.913(39)	&	&9.67(21)	&-0.792(28)	&0.082(85)	&0.050\\
		\hline
	\end{tabular}
\end{table*}
% source year variation: fit_out_yearly_gradients_table_b_stats.txt

@@

%crossing distances
\textbf{The mean and median velocity fit exponents acquired from the Helios data} are very similar, which indicates that they just as well can be kept identical so that the basic shape of the frequency distribution does not change with distance. Contrary, the mean and median fits for the magnetic field strength cross each other at \SI{0.339}{\au} \textbf{(see Table~\ref{tab:mean_median_fit_parameter})} and the mean \hl{is \textbf{getting slightly} lower than the median at smaller distances }. Thus, below that distance the \textbf{frequency distribution} cannot well be described anymore by a lognormal function\textbf{, because the mean of a lognormal function has to be larger than its median (as mentioned in Sect.~\ref{sec:frequency_distribution}), that is, the location of the crossing indicates that the parameter's distribution is not anymore of a lognormal shape thereafter.} The fits for the proton temperature show a similar behavior, having an \textbf{extrapolated} intersection at \SI{0.082}{\au}. Therefore the extrapolation of the magnetic field and temperature distribution frequencies to the PSP orbit by applying lognormal functions is limited. \textbf{The crossing points limit the regions where the distribution's shapes can still be considered lognormal.}

\textbf{In order to still fit and extrapolate lognormal functions with the data, we assume that the shapes can be considered lognormal at all distances. For the coming frequency distribution fit function we reduce the fit exponents $e_\text{med}$ and $e_\text{avg}$ to only one.} It should be noted that this simplification leads to slightly larger modeling errors, especially in case of the magnetic field strength.
